---
title: "Do heavier penguins have longer flippers? An R analysis"
author: "PAROMITA SENGUPTA"
date: "2026-09-28"
categories: [R, penguins, data analysis]
---

## Introduction

Penguins on the Palmer Archipelago come in three species that look quite
different in size. In this post I use R to ask two simple questions:

1. How does body mass differ between the three species?
2. Within a species, do penguins with longer flippers weigh more?

**Data:** [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
Palmer Station Antarctica LTER. The data ship inside the `palmerpenguins` R
package, so nothing is downloaded when this page is built. The data are
released under the [CC0 licence](https://creativecommons.org/publicdomain/zero/1.0/).

Citation: Horst AM, Hill AP, Gorman KB (2020). *palmerpenguins: Palmer
Archipelago (Antarctica) penguin data.* R package version 0.1.0.
<https://doi.org/10.5281/zenodo.3960218>. The measurements were collected by
Gorman KB, Williams TD, Fraser WR (2014), *PLoS ONE* 9(3): e90081.

## Setup

The setup chunk loads packages. I silence package start-up messages and
warnings on purpose, because they add noise and no information.

```{r}
#| label: setup
#| message: false
#| warning: false
library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(knitr)
```

## Looking at the data

```{r}
#| label: load-data
glimpse(penguins)
```

Some birds have missing measurements. I count them per column before deciding
what to do.

```{r}
#| label: missing-values
penguins |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  select(species, body_mass_g, flipper_length_mm, sex)
```

```{r}
#| label: clean-data
penguins_clean <- penguins |>
  filter(!is.na(body_mass_g), !is.na(flipper_length_mm))

n_dropped <- nrow(penguins) - nrow(penguins_clean)
```

I dropped `{r} n_dropped` birds that lacked a body mass or flipper length,
leaving `{r} nrow(penguins_clean)` penguins.

## Body mass by species

```{r}
#| label: tbl-mass-summary
#| tbl-cap: "Body mass (g) summarised by species."
mass_summary <- penguins_clean |>
  group_by(species) |>
  summarise(
    n = n(),
    mean_mass = round(mean(body_mass_g)),
    sd_mass = round(sd(body_mass_g)),
    min_mass = min(body_mass_g),
    max_mass = max(body_mass_g)
  )

kable(mass_summary)
```

```{r}
#| label: fig-penguin-mass
#| fig-cap: "Body mass of penguins by species."
ggplot(penguins_clean, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = "Species", y = "Body mass (g)") +
  theme_minimal()
```

@fig-penguin-mass and @tbl-mass-summary show that Gentoo penguins are the
heaviest, with a mean of
`{r} mass_summary$mean_mass[mass_summary$species == "Gentoo"]` g. Adelie and
Chinstrap penguins are much lighter and overlap heavily with each other.

## Flipper length and body mass

```{r}
#| label: fig-flipper-mass
#| fig-cap: "Flipper length against body mass. Each species has its own fitted line."
ggplot(penguins_clean,
       aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(x = "Flipper length (mm)", y = "Body mass (g)", colour = "Species") +
  theme_minimal()
```

To put a number on the trend, I fit a linear model of body mass on flipper
length, with species as an additional predictor.

```{r}
#| label: linear-model
model <- lm(body_mass_g ~ flipper_length_mm + species, data = penguins_clean)
kable(round(summary(model)$coefficients, 2))
```

## What I found

- Gentoo penguins are clearly heavier than the other two species.
- Within the data, longer flippers go with heavier birds. The model estimates
  about `{r} round(coef(model)["flipper_length_mm"], 1)` g more body mass for
  each extra millimetre of flipper, holding species fixed.
- This is an observational association. It says nothing about whether flipper
  length causes weight.